---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    fig_height: 4
    highlight: pygments
    theme: spacelab
---


## Load packages

```{r load-packages, message = FALSE, echo = TRUE}
library(ggplot2)
library(dplyr)
```

## Loading and preprocessing the data


```{r load-data, echo = TRUE}
activity <- read.csv("activity.csv", stringsAsFactors=FALSE)
str(activity)
```

### Summarizing steps per day 

```{r summarice by steps per day ignore na, echo = TRUE}
activity_steps_day <- na.omit(activity) %>% 
  group_by(date) %>%
  summarise(steps_day=sum(steps))

head(activity_steps_day)
tail(activity_steps_day)
```

### Summarizing steps on average through the 24 hours

```{r summarice by steps per hour ignore na, echo = TRUE}
activity_steps_hour <- na.omit(activity) %>% 
  group_by(interval) %>%
  summarise(steps_interval=mean(steps))

activity_steps_hour <- activity_steps_hour %>% mutate(hour = trunc(interval/100))

```

* * *

## What is the mean total number of steps taken per day?

For this part of the assignment the missing values in the dataset are ignore.


### Histogram of the total number of steps taken each day


```{r  echo = TRUE}
ggplot(data=activity_steps_day, aes(activity_steps_day$steps_day)) + 
  geom_histogram(col="red", 
                 aes(fill=..count..)) +
  scale_fill_gradient("Count", low = "red", high = "green") +
  labs(title="Histogram steps per day") +
  labs(x="Steps", y="Count Days") 

```


### Mean and median of total number of steps taken per day


```{r mean and median, echo = TRUE}
activity_steps_day %>% 
  summarise(steps_median=median(steps_day))

activity_steps_day %>% 
  summarise(steps_mean=mean(steps_day))

```



## Average daily activity pattern


### Time series plot 5 minute interval


```{r echo = TRUE}
ggplot(data=activity_steps_hour, aes(x=interval, y=steps_interval)) + 
  geom_bar(stat="identity") + 
  labs(title="Histogram steps taken on average per interval") +
  labs(x="5 min Interval", y="Steps") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


**From the histogram we can observe that the most active hour is 8AM.


```{r echo = TRUE}
ggplot(data=activity_steps_hour, aes(x=interval, y=steps_interval)) + geom_line() +
  labs(title="Time series steps taken on average per interval") +
  labs(x="5 min Interval", y="Steps") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


### 5-minute interval which contains on average maximum number of steps across all the days in the dataset



```{r max number of steps on average, echo = TRUE}
activity_steps_hour %>% 
  summarise(steps_max=max(steps_interval))

```

###The 5-minute interval with more steps on average is at 8:35am.  I could find the max average but I was not able to display in the same result which time specific interval was it.  I look for it manually in the data.

* * *

## Imputing missing values

### Total number of missing values in the dataset


```{r echo = TRUE}
sum(is.na(activity$steps))

```

Total number of rows with missing values in the data set is 2304.


### strategy for filling missing values is assiging the 5-minute interval mean:

```{r echo = TRUE}
activity_steps_hour %>% 
  summarise(steps_mean=mean(steps_interval))

```



```{r echo = TRUE}
activity2 <- activity
activity2$steps[is.na(activity2$steps)] <- 37

sum(is.na(activity2$steps))
```
 

### Histogram of the total number of steps taken each day (Missing values filled)


```{r  echo = TRUE}

activity_steps_day <- activity2 %>% 
  group_by(date) %>%
  summarise(steps_day=sum(steps))

```


```{r  echo = TRUE}

ggplot(data=activity_steps_day, aes(activity_steps_day$steps_day)) + 
  geom_histogram(col="red", 
                 aes(fill=..count..)) +
  scale_fill_gradient("Count", low = "red", high = "green") +
  labs(title="Histogram steps per day") +
  labs(x="Steps", y="Count Days") 

```


### Mean and median of total number of steps taken per day (Missing values filled)


```{r echo = TRUE}
activity_steps_day %>% 
  summarise(steps_median=median(steps_day))

activity_steps_day %>% 
  summarise(steps_mean=mean(steps_day))

```


**Before filling missing values the mean and the median were almost equal (median=10765, mean=70766)  After filling the missing values the Mean and median has changed, the mean is now greater than median by small difference.**



* * *

## Pattern differences in activity between weekdays and weekends


### define day of the week in the dataset and recreate the average 5-minute interval dataset


```{r echo=TRUE}

activity2$day <- weekdays(as.Date(activity$date))

activity2$wday <- "weekday"
activity2$wday[activity2$day=="Saturday"] <- "weekend"
activity2$wday[activity2$day=="Sunday"] <- "weekend"


activity_steps_hour <- activity2 %>% 
  group_by(interval, wday) %>%
  summarise(steps_interval=mean(steps))

```


### panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 


```{r echo=TRUE}

ggplot(data=activity_steps_hour, aes(x=interval, y=steps_interval)) + geom_line() +
  labs(title="Time series steps taken on average per interval") +
  labs(x="5 min Interval", y="Steps") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~ wday)

```



